The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-04-01 Paper.Rmd
edits to Discussion

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., local pest abundance
time series) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in an Appendix showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results show that, for many
ecological phenomena, a complete understanding is impossible without 
characterizing dependence between variables using copulas.

<!--8) A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.-->
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
convenient computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists need only to become aware of these tools
and understand them conceptually to start applying them broadly to ecological
problems. We created a variety of interrelated surrogate (randomization) procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful in ecological applications.

>>>>

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., average regional pest abundance computed from local pest abundances) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in Appendix \ref{SM-Consequence_TL_simulation} 
showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results demonstrate that, for many
ecological phenomena, complete understanding is impossible without 
characterizing dependence between variables using copulas.

<!--8) A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.-->
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists need only to become aware of these tools
and understand them conceptually to start applying them broadly. 
We created a variety of interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecological applications.
